From Reversible Programs to Univalent Universes and Back

### 2017-05-03 upi.tex

The early developments of reversible programming languages started with a
conventional programming language, e.g., an extended λ-calculus, and either
extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible} or
imposed constraints on the conventional control flow constructs to make them
reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. More modern approaches
recognize that reversible programming requires a fresh approach and should be
designed from first principles without the detour via conventional irreversible
languages~\cite{Yokoyama:2008:PRP, Mu:2004:ILRC, abramsky2005structural,
  DiPierro:2006:RCL:1166042.1166047}.

@@

>>>>

The early developments of reversible programming languages started with a
conventional programming language, e.g., extended $\lambda$-calculi, and either
(i) extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible} or
(ii) imposed constraints on the control flow constructs to make them
reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. More modern approaches
recognize that reversible programming languages require a fresh approach and
should be designed from first principles without the detour via conventional
irreversible languages~\cite{Yokoyama:2008:PRP, Mu:2004:ILRC,
  abramsky2005structural, DiPierro:2006:RCL:1166042.1166047}.

%%%%%
\subsection{The $\Pi$ Family of Languages}

In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
the~$\Pi$ family of reversible languages based on type isomorphisms and
commutative semiring identities. The fragment without recursive types is
universal for reversible boolean
circuits~\citep{James:2012:IE:2103656.2103667}. With the addition of recursive
types and trace operators~\citep{Hasegawa:1997:RCS:645893.671607}, the variant
of $\Pi$ becomes a Turing-complete reversible
language~\citep{James:2012:IE:2103656.2103667,rc2011}. Before identifying a
``featherweight'' representative to use for the formal development in this
paper, we illustrate the main underlying ideas for the entire family of
languages using some examples.

\amr{EXAMPLES OF CIRCUITS}

Although writing circuits using the raw syntax for combinators is tedious, the
examples illustrate the programming language nature of this family of $\Pi$
languages. In other work, one can find a compiler from a conventional functional
language to circuits~\citep{James:2012:IE:2103656.2103667}, a systematic
technique to translate abstract machines to $\Pi$~\citep{rc2012}, and a
Haskell-like surface language~\citep{theseus} which can ease writing
circuits. All that reinforces the claim that this is a practical approach to
reversible programming.

\amr{EXAMPLES OF OPTIMIZATIONS}

@@

\end{verbatim}

%%%%%
\subsection{$\Pi_{\mathbb{2}}$}

\begin{verbatim}
